---
title: "Introduction to RNAseq"
author: Nicole Black
date: 07/24/22
---

```{r setup}
    knitr::opts_knit$set(root.dir = normalizePath("/Users/brandonye/Desktop/fa22/QBIO 490/fa22_qbio_490_brandon/analysis_data")) 
```

```{r}
# install, only if needed, and load packages (SummarizedExperiment and TCGAbiolinks)
install.packages(SummarizedExperiment)
library(SummarizedExperiment)

install.packages(TCGAbiolinks)
library(TCGAbiolinks)
```

Make sure you have access to rna_se. If it is not already in your workspace (ie already queried, downloaded, and prepared), let Nicole know NOW so we can get you the data a different way.

###### (1) Summarized Experiment Basics

Let's take a look at rna_se (RNA Summarized Experiment). Call head() and str(). Also explore the data frame by double clicking on it from the Environment panel (or calling "View(rna_se)").

```{r Exercise 1.1}
# write code here
head(rna_se)
str(rna_se)
```

Use the dim() function to find the dimensions of the colData, rowRanges, and unstranded dataframes. Record them below:

  ~ colData (rows, cols): 1226, 87
  ~ rowRanges: 60660, 10
  ~ unstranded: 60660, 1226

```{r Exercise 1.2}
# write code here
dim(rna_se@colData)
dim(rna_se@rowRanges@elementMetadata)
dim(rna_se@assays@data$unstranded)
```

Run the following code and fill out a description of the rows/columns of our data frames.You can use this as a cheat sheet moving forward.
  ~ colData
        - Rows: patients
        - Columns: info about patients
  ~ rowRanges
        - Rows: genes
        - Columns: info about genes
  ~ unstranded (in assays)
        - Rows: genes
        - Columns: patients
  
```{r Exercise 1.3}
# colData: 
rownames()
colnames()

# rowRanges:
str()
# what are the rows... use your logic

# for unstranded, don't call anything, just use your logic from exercise 1.2 and and think it through!
```

###### (2) ColData

View the first 15 values in the age_at_index column.

70 64 54 80 68 84 52 49 40 50 46 62 46 69 78

```{r Exercise 2.1}
# write code here
rna_se@colData$age_at_index[1:15]
```

Call summary on the age_at_index column. Record the output below:
    
  Min. 1st Qu.  Median    Mean 3rd Qu.    Max.    NA's 
  26.00   48.00   58.00   58.27   67.00   90.00       1 
    
```{r Exercise 2.2}
# write code here
summary(rna_se@colData$age_at_index)
```

Create a data frame called rna_clinical that contains data from colData, except for the NA patient and their data.

```{r Exercise 2.3}
age_mask <-  ifelse(is.na(rna_se@colData$age_at_index), FALSE, TRUE)
rna_clinical <-  rna_se@colData[age_mask, ]

summary(rna_clinical) # NA's removed

rna_clinical <- rna_se@colData[!is.na(rna_se@colData$age_at_index), ] 
```

Use the as.data.frame() function to change this from an S4 object to a data frame object.

```{r Exercise 2.4}
rna_clinical <- as.data.frame(rna_clinical)
```

Make a mask and subset out the treatments column.

```{r Exercise 2.5}
treatments_mask <- ifelse(colnames(rna_clinical) == "treatments" | colnames(rna_clinical) == "disease_type" | colnames(rna_clinical) == "primary_type", FALSE, TRUE)
rna_clinical <- rna_clinical[,treatments_mask] 

sum(!treatments_mask)

```

Save rna_clinical to your local machine

```{r Exercise 2.6}
write.csv(rna_clinical, "/Users/brandonye/Desktop/fa22/QBIO 490/fa22_qbio_490_brandon/analysis_data/brca_rna_clincial_data.csv", row.names = TRUE)
```

###### (3) RowRanges

Call head() on rowRanges@elementMetadata:

```{r Exercise 3.1}
# write code here
head(rna_se@rowRanges@elementMetadata)
```

Create a data frame called rna_genes that contains data from elementMetadata.

```{r Exercise 3.2}
rna_genes <- rna_se@rowRanges@elementMetadata
```

Use the as.data.frame() function to change this from an S4 object to a data frame object

```{r Exercise 3.3}
rna_genes <- as.data.frame(rna_genes)
```

Save rna_genes to your local machine.

```{r Exercise 3.4}
write.csv(rna_genes, "/Users/brandonye/Desktop/fa22/QBIO 490/fa22_qbio_490_brandon/analysis_data/brca_rna_gene_data.csv", row.names = FALSE)
```

###### (4) Assays

What are the names of the 6 assays?

```{r Exercise 4.1}
# access the assays here
rna_se@assays
```

Create a new data frame called "rna_counts" that contains the data from unstranded. Recall the patient with NA age data? We need to exclude that patient. We can use the same mask to do this.

```{r Exercise 4.2}
rna_counts <- rna_se@assays@data$unstranded
rna_counts_mask <- ifelse(is.na(rna_se@colData$age_at_index), F, T)

rna_counts = rna_counts[,rna_counts_mask]
rna_counts <- as.data.frame(rna_counts)

rna_counts <- rna_se@assays@data$unstranded[, !age_mask]
```

Fetch the data for the first 5 patients and the first 15 rows.

```{r Exercise 4.3}
# write code here
rna_counts[1:15,1:5]
```

Save rna_counts to your local machine (warning: this will likely take a while!)

```{r Exercise 4.4}
write.csv(rna_counts, "/Users/brandonye/Desktop/fa22/QBIO 490/fa22_qbio_490_brandon/analysis_data/brca_rna_count_data.csv", row.names = TRUE)
```
