---
title: "Intro to MAF - Part II"
author: Nicole Black
date: 9/27/22
---

###### (1) Set Up

Set your path:

```{r setup}
    knitr::opts_knit$set(root.dir = normalizePath("/Users/brandonye/Desktop/fa22/QBIO 490/fa22_qbio_490_brandon/analysis_data/brca_clinical_data.csv")) 
```

Load in necessary packages:

```{r}
# load maftools
library(TCGAbiolinks)
library(maftools)
```

Read in the clinical data.

```{r Exercise 1.1}
clinical <- read.csv("/Users/brandonye/Desktop/fa22/QBIO 490/fa22_qbio_490_brandon/analysis_data/brca_clinical_data.csv")
```

Read in your maf object.

```{r Exercise 1.2}
maf_query <- GDCquery(
    project = "TCGA-BRCA", 
    data.category = "Simple Nucleotide Variation", 
    access = "open", # we only have access to somatic mutations which are open access
    data.type = "Masked Somatic Mutation", 
    workflow.type = "Aliquot Ensemble Somatic Variant Merging and Masking"
)

GDCdownload(maf_query) ## only run this if you haven't already!! (if you know you've run it but are still getting an issue, check your working directory file path and the file path for where you saved GDCdata)

maf <- GDCprepare(maf_query)
maf_object <- read.maf(maf = maf, 
                       clinicalData = clinical, ## make sure you changed the name of "bcr_patient_barcode" to "Tumor_Sample_Barcode" (see exercise 1.2 of Intro_to_MAF.Rmd)
                       isTCGA = TRUE)
```

###### (2) Fisher's Exact Test - Demo

Rewrite the vital_status and breast_carcinoma_estrogen_receptor_status columns of your clinical df as factors.

```{r}
clinical$vital_status <- factor(clinical$vital_status)

clinical$breast_carcinoma_estrogen_receptor_status <- factor(clinical$breast_carcinoma_estrogen_receptor_status, levels = c("Positive", "Negative"))
```

Create a contingency table using table().

```{r}
contig <- table(clinical$breast_carcinoma_estrogen_receptor_status, clinical$vital_status)
```

Visualize our contingency table using a mosaic plot.

```{r}
mosaicplot(contig)
```

Run our Fisher's exact test.

```{r}
fisher_test <- fisher.test(contig)
```

View the results of our test.

```{r}
fisher_test

fisher_test$p.value
```

What does this p-value tell us?

Does this make sense given the biology of cancer?

###### (3) Mutational Co-Occurence Example

Subset your maf_object to create geneA_maf and geneB_maf.

```{r}
## half class do: TP53 + PIK3CA
## other half do: MUC16 + RYR2

geneA_maf <- subsetMaf(maf = maf_object,
                       genes = "MUC16") ## fill in with your gene name as a string

geneB_maf <- subsetMaf(maf = maf_object,
                       genes = "RYR2")
```

Now let's start to build our contingency table.

Access the barcodes of the patients with mutations in genes A and the barcodes of the patinets with mutations in gene B. HINT: Boolean indexing! How do we access patient barcodes within the MAF object without duplicate patients? @data or @clinical.data?

```{r}
mut_pats_geneA = geneA_maf@clinical.data$Tumor_Sample_Barcode
mut_pats_geneB = geneB_maf@clinical.data$Tumor_Sample_Barcode
```

Get the lengths of these two vectors.

```{r}
num_pats_geneA = length(mut_pats_geneA)
num_pats_geneB = length(mut_pats_geneB)
```

Use intersect() to get the patients with mutations in both gene A AND gene B. Then find the length of that vector. What does this tell us in our contingency table?

```{r}
mut_pats_geneAB = intersect(mut_pats_geneA, mut_pats_geneB)
num_AB = length(mut_pats_geneAB)
```

Now, determine the number of patients with one mutation and not the other. HINT: Do some subtraction with your above vectors.

```{r}
num_A_only = num_pats_geneA-num_AB
num_B_only = num_pats_geneB-num_AB
```

Finally, find the number of patients without either mutation. HINT: We know the total number of patients (based on our original maf_object), and we just calculated the number of patients with mutations (A only + B only + AB).

```{r}
num_neither = nrow(maf_object@clinical.data)-(num_A_only+num_B_only+num_AB)
```

Create your contingency table:

```{r}
contig <- matrix(c(num_AB, 
                         num_B_only,
                         num_A_only,
                         num_neither), 
                       nrow=2)

# view the contingency table
contig

# view the contingency table as a mosaic
mosaicplot(contig)
```

Run your Fisher exact test!

```{r}
fisher_test <- fisher.test(contig)
fisher_test
```

Are your mutations co-occurent or mutually exclusive?

###### (4) Maftools somaticInteractions()

Luckily, we don't have to do the above work by hand for each gene pair we want to look at because maftools has a function that does it for us!

Run somaticInteractions for your above genes by passing in a vector with your gene names into the genes argument.

```{r}
somaticInteractions(maf = maf_object,
                    genes = c("MUC16", "RYR2"),
                    pvalue = c(0.05, 0.1)) ## these are the default p-value significance levels

## notice the $event column tells us if our mutations are co-occurent or mutually exclusive!
```

Run the following line of code for the top 25 genes.

```{r}
somaticInteractions(maf = maf_object,
                    top = 25,
                    pvalue = c(0.05, 0.1))
```

Pick a gene pair that is either significantly co-occurant or significantly mutually exclusive and look up the genes! Does this make sense? Do you see any literature that expresses similar/oppositional views on these genes?

